MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-17 results.tex
added RSCC reference
Added reference to Table S2 in 4.2

\section{Results}
In the Top8000 system there are a total of 235 named rotamers which reach above the 0.3\% outlier cutoff and include $\ge$ 4 datapoints. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see section \ref{sec_md:chi_stats}), which improves on both the modal and the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S2 describes the frequency and count of each rotamer in the given residue type and Table S3 has central dihedral and covalent angle values for each rotamer. The latter numbers are available in CSV format on GitHub under \texttt{datasets/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/BJH_Dissertation}. The more useful multi-dimensional distributions are also on GitHub, as linked in section \ref{sec_md:contours}.

@@

For validation purposes, users are often most interested in the binary issue of whether a sidechain conformation is rotameric or an outlier, i.e. falls within or outside the rotamer distribution. With the Top500 data, contours were smooth and reproducible (between different datapoint selections) only out to the 1\% level. Due to greater numbers and increased reliability in the filtered Top8000 distributions, a lower outlier cutoff is now feasible as well as desirable, since an even lower percentage of the reference datapoints are now dubious. We also preferred to keep outlier numbers for unfiltered, general data roughly equal between the previous and new systems. Therefore a count of the 1\% outliers by residue type in the unfiltered dataset was performed using the Top500 contours, and those numbers were compared to outlier counts at several different Top8000 contour cutoffs. A new cutoff of 0.3\% matched best, was found to behave smoothly, and happens also to match the 3$\sigma$ level for a normal distribution (3 out of 1000). Further, along with favored and outlier, a new category has now been added: \textit{allowed}, for scores $\ge$0.3\% and $<$2.0\%. This new category lets users know if a given conformation is at the edge of the given rotamer's distribution, close to the outlier region. This change matches the 3-part system, and the division at 2.0\%, long judged useful for Ramachandran criteria \cite{Kleywegt1996, Hooft1996, Lovell:2003uq}.

An important question is how the new system will change rotamer analysis, specifically the number and identity of outliers. A rotamer outlier is simply a conformation that lies outside the outlier contours of the reference dataset, in this case either the Top500 or the Top8000. For residues that have low-energy sidechain conformations, nothing will change. However, a sidechain conformer in a strained position near the outlier edge could change status between the Top500 and Top8000 analyses. Figure \ref{fig:contour_compare} shows Top500 and Top8000 allowed regions superimposed, for three residue types. For the most part, the contours overlap, but with some differences near the edges. Asp shows very little change, Trp has a significantly larger allowed region now than in the Top500 system, and the Ile Top8000 contours have tightened up relative to the Top500. Thus, differences are only in low-population regions of $\chi$ space, and overall outlier counts will be nearly identical.

>>>>

\section{Results}
In the Top8000 system there are a total of 235 named rotamers which reach above the 0.3\% outlier cutoff and include $\ge$ 4 datapoints. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see section \ref{sec_md:chi_stats}), which improves on both the modal and the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S3 describes the frequency and count of each rotamer in each residue type and the tables in Section S5 have central dihedral and covalent angle values for each rotamer. The latter numbers are available in CSV format on GitHub under \texttt{datasets/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/BJH_Dissertation}. The more useful multi-dimensional distributions are also on GitHub, as linked in section \ref{sec_md:contours}.

@@

For validation purposes, users are often most interested in the binary issue of whether a sidechain conformation is rotameric or an outlier, i.e. falls within or outside the rotamer distribution. With the Top500 data, contours were smooth and reproducible (between different datapoint selections) only out to the 1\% level. Due to greater numbers and increased reliability in the filtered Top8000 distributions, a lower outlier cutoff is now feasible as well as desirable, since an even lower percentage of the reference datapoints are now dubious. We also preferred to keep outlier numbers for unfiltered, general data roughly equal between the previous and new systems. Therefore a count of the 1\% outliers by residue type in the unfiltered dataset was performed using the Top500 contours, and those numbers were compared to outlier counts at several different Top8000 contour cutoffs. A new cutoff of 0.3\% matched best, was found to behave smoothly, and happens also to match the 3$\sigma$ level for a normal distribution (3 out of 1000). \textcolor{red}{Table S2 reports the outlier counts in the unfiltered dataset for both the Top500 and Top8000 contours.} Further, along with favored and outlier, a new category has now been added: \textit{allowed}, for scores $\ge$0.3\% and $<$2.0\%. This new category lets users know if a given conformation is at the edge of the given rotamer's distribution, close to the outlier region. This change matches the 3-part system, and the division at 2.0\%, long judged useful for Ramachandran criteria \cite{Kleywegt1996, Hooft1996, Lovell:2003uq}.

An important question is how the new system will change rotamer analysis, specifically the number and identity of outliers. A rotamer outlier is simply a conformation that lies outside the outlier contours of the reference dataset, in this case either the Top500 or the Top8000. For residues that have low-energy sidechain conformations, nothing will change. However, a sidechain conformer in a strained position near the outlier edge could change status between the Top500 and Top8000 analyses. Figure\textcolor{red}{s} \ref{fig:contour_compare} \textcolor{red}{and S1 show} Top500 and Top8000 allowed regions superimposed for \textcolor{red}{all residues with two $\chi$s}. For the most part, the contours overlap, but with some differences near the edges. Asp shows very little change, Trp has a significantly larger allowed region now than in the Top500 system, and the Ile Top8000 contours have tightened up relative to the Top500. Thus, differences are only in low-population regions of $\chi$ space, and overall outlier counts will be nearly identical.
